---
title: "Machine Learning Assignment"
author: "Stephen Checkley"
date: "June 2015"
output: html_document
---

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. 

6 participants were asked to perform barbell lifts corrects and incorrectly in 5 different ways (A, B, C, D, and E) and data was collected from accelerometers on the belt, forearm, arm, and dumbbell of the participants.

The goal of this project is to predict the manner in which the exercise was performed.

#Obtaining the Exercise Data
```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(curl)
library(caret)
library(rpart)
library(rpart.plot)
require(randomForest)
library(gplots)
library(RColorBrewer)
```

```{r}
#import the data from the web
testing_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
training_url <-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"

if (!file.exists("pml-testing.csv")) {
  download.file(testing_url, destfile="pml-testing.csv", method="curl")
}

if (!file.exists("pml-training.csv")) {
  download.file(training_url, destfile="pml-training.csv", method="curl")
}

```

#Importing and Cleaning the Data
```{r}
#load the data
data.testing <- read.csv("pml-testing.csv",na.strings=c("NA", "#DIV/0!"))
data.training <- read.csv("pml-training.csv",na.strings=c("NA", "#DIV/0!"))
#View(data.training)

#remove missing values
data.testing <- data.testing[, colSums(is.na(data.testing)) == 0] 
data.training <- data.training[, colSums(is.na(data.training)) == 0] 

#remove unnecessary columns
data.testing <- data.testing[ , !names(data.testing) %in% c("X", "user_name", "raw_timestamp_part_1", "raw_timestamp_part_2", "cvtd_timestamp", "new_window","num_window")]
data.training <- data.training[ , !names(data.training) %in% c("X", "user_name", "raw_timestamp_part_1", "raw_timestamp_part_2", "cvtd_timestamp", "new_window","num_window")]
data.testing$problem_id <- NULL
```

#Exploratory Data Analysis
```{r}
data4map <- cor(data.training[, names(data.training) != "classe"])
my_palette <- colorRampPalette(c("darkblue", "blue", "white", "darkred", "red"))(n = 1000)
heatmap.2(data4map,
          col = my_palette,
          dendrogram="both",
          density.info = "none",
          trace="none",
          margins = c(5,10),
          key=TRUE,
          scale="none"
          )
```

The heat-map does not show any clear correlation between the variables in the data set and the dendrograms do not show any obvious clustering.

#Preparation for Modelling
##Data Preparation
```{r}
inTrain <- createDataPartition(y = data.training$classe, p = 0.7, list = FALSE)
data4training <- data.training[inTrain, ] #training subset
data4testing <- data.training[-inTrain, ] #testing subset
```

The training data set is split with 70% allocated for training and 30% allocated to model testing.

##Model Preparation
```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(doMC)
```
```{r}
registerDoMC(cores = detectCores(logical=FALSE)) #set up parallel workers
set.seed(8418)
trControl <- trainControl(method = "cv", number = 10, allowParallel = TRUE, verboseIter = FALSE)
```
10-fold Cross validation is set, paralellized across all available CPU cores.

```{r}
modfit <- train(classe ~ ., data=data4training, method="rf", trControl=trControl, prox=FALSE)
modfit
```
Machine learning is applied to the training data subset using the Random Forest algorithm. The resulting model fit is reported to be 99% accurate using 52 predictors over 13737 samples (see Appendix for a map of the final model tree and details of variable importance).

##Model Validation
```{r}
modfit.test <- predict(modfit, newdata = data4testing)
cfMatrix <- confusionMatrix(data4testing$classe, modfit.test)
cfMatrix

missClass = function(values, model.test) {
  sum(model.test != values)/length(values)
}
error = missClass(data4testing$classe, modfit.test)
error
```
The model is validated using the training data subset, outputting a confusion matrix to summarize the results of the test. The model reports 99.1% accuracy with 0.9% out of sample error.

#Model Predictions
Using the real testing data set the model predicts the following exercise classes from 20 observations:
```{r}
test.result <- predict(modfit, data.testing)
test.result
```

#Appendix
```{r, echo=FALSE, warning=FALSE, message=FALSE}
tree <- rpart(classe ~ ., data=data4training, method="class")
```
```{r}
prp(tree)
```

Diagrammatic representation of the final model tree.

```{r}
varImp(modfit)
```
Analysis of variable importance for the model.